Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations

### 2022-01-31 document/source/body.tex
proof read GC

Non-deterministic simulation may have non-zero variance in, for example, actor positions, which may result in intermittent assertion failure and failing to detect bugs and leading to false confidence in the system. When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers. It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience. However, the situation is very different for AV development and testing. Thus, our main research questions are:
{\em How can one assess whether a simulation environment is deterministic?} and 
{\em How can one determine and control the simulation variance?}

@@

There are numerous game engines with their associated development environments that could be considered suitable for AV development, e.g.\ Unreal Engine
, Unity, CryEngine.\footnote{\url{https://www.cryengine.com/}} 
% ~\cite{CryEngine_main_website}
Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{CARLA_paper}, AirSim\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud-based simulation, e.g.\ Nvidia Drive Constellation\footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
%~\cite{UE4_main_website}
% ~\cite{Unity_main_website}
%~\cite{carla_main_website}
% ~\cite{AirSim_main_website}
% ~\cite{Apollo_main_website}
% ~\cite{nvidia_constellation}

@@

The physics engine operates according to a time step, $dt$. The shorter this time step is, the smoother the interpretation of the physical dynamics will be. To use a fixed physics time step, the user's display refresh rate needs to be known in advance. This requires an update loop to take less than one render tick (one frame of real world time). Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$. However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations\footnote{\url{https://gafferongames.com/post/fix_your_timestep/}}.%~\cite{gaffer}. 
%
Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allow computational headroom for slower hardware. Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine~\cite{UE4_substepping}. If the engine tries to render between physics updates, \textit{residual lag} can occur, which may result in frames arriving with a delay to the simulated physics. Thus, extrapolation between frames may need to be performed to smooth transition between scenes. Note that both residual lag and extrapolation could affect perception stack testing. In exceptional cases, where computational resources are scarce, the fixed time step can be greater than the time between render ticks and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming.

@@

% NEW TEXT V3
It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which limited by the fixed bit width available and potentially resulting rounding errors~\cite{FloatingPointsBook,goldberg1991every}. As a consequence, floating-point arithmetic is not associative~\cite{Kapre2007} and depends on execution order where some authors even suggest avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}. 
%
In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions but we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. So, even if the result of a operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}. 
%
However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, 
parallelisation within the runtime environment or at the hardware level and if the execution is performed on a GPU rather than a CPU (which may have different register widths~\cite{Whitehead2011}) may all affect the execution order. 
%
In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable, hardware, configuration and execution order.

@@

It is unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should give the same output. However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} 
% ~\cite{unity_multithreading} 
or if the scheduler simply randomly selects from a set of threads with equal priority, this may alter an otherwise deterministic sequence of events.

Similar to thread scheduling, scheduling at the hardware level on a multi-core system determines on which processor core to execute processes. This may be decided based on factors such as throughput, latency or CPU utilisation. Scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, or if executing across parallel GPU architecture, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution. Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads, thus promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.

>>>>

Non-deterministic simulation may have non-zero variance in, for example, actor positions, which may result in intermittent assertion failure, failing to detect bugs and potentially leading to false confidence in the system under test. When used for gaming, game engines do not need to be deterministic nor do they have any requirements on the limits of permissible variance; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers. It could even be argued that simulation variance is a feature that enhances gaming and improves the user experience. However, the situation is very different for AV development and testing. Thus, our main research questions are:
{\em How can one assess whether a simulation environment is deterministic?} and 
{\em How can one determine and control the simulation variance?}

@@

There are numerous game engines with their associated development environments that could be considered suitable for AV development, e.g.\ Unreal Engine
, Unity\footnote{\url{https://unity.com/}} and CryEngine.\footnote{\url{https://www.cryengine.com/}} 
% ~\cite{CryEngine_main_website}
Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{CARLA_paper}, AirSim\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud-based simulation, e.g.\ Nvidia Drive Constellation\footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.
%~\cite{UE4_main_website}
% ~\cite{Unity_main_website}
%~\cite{carla_main_website}
% ~\cite{AirSim_main_website}
% ~\cite{Apollo_main_website}
% ~\cite{nvidia_constellation}

@@

The physics engine operates according to a time step, $dt$. The shorter this time step is, the smoother the interpretation of the physical dynamics will be. To use a fixed physics time step, the user's display refresh rate needs to be known in advance. This requires an update loop to take less than one render tick (one frame of real world time). Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$. However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations\footnote{\url{https://gafferongames.com/post/fix_your_timestep/}}. %~\cite{gaffer}. 
%
Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allow computational headroom for slower hardware. Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine~\cite{UE4_substepping}. If the engine tries to render between physics updates, \textit{residual lag} can occur, which may result in frames arriving with a delay to the simulated physics. Thus, extrapolation between frames may need to be performed to smooth transition between scenes. Note that both residual lag and extrapolation could affect perception stack testing. In exceptional cases, where computational resources are scarce, the fixed time step can be greater than the time between render ticks and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming.

@@

% NEW TEXT V3
It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which is limited by the fixed bit width available and can potentially result rounding errors~\cite{FloatingPointsBook,goldberg1991every}. As a consequence, floating-point arithmetic is not associative~\cite{Kapre2007} and depends on execution order where some authors even suggest avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}. 
%
In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions but we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. So, even if the result of a mathematical operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} when repeated for implementations that conform to the IEEE floating-point standard~\cite{8766229}. 
%
However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, 
parallelisation within the runtime environment or at the hardware level and if the execution is performed on a GPU rather than a CPU\footnote{ CPU and GPU processors may have different register widths~\cite{Whitehead2011} } may all affect the execution order. 
%
In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable, hardware, configuration and execution order.

@@

It is unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should therefore give the same output. However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} 
% ~\cite{unity_multithreading} 
or if the scheduler simply randomly selects from a set of threads with equal priority, this may alter an otherwise deterministic sequence of events.

Similar to thread scheduling, scheduling at the hardware level on a multi-core system determines on which processor core to execute processes. This may be decided based on factors such as throughput, latency or CPU utilisation. Scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, or executing across a parallel GPU architecture, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution. Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads, thus promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
